\pkg{Rgbp}: An R Package for Hierarchical Modeling and Method Checking

### 2013-06-20 jss/article.tex
typos

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Joseph Kelly, Carl Morris, Hyung Suk Tak} %% comma-separated
\Plaintitle{Rgbp: Bayesian Hierarchical Modeling and Frequentist Method Check} %% without formatting

@@

where $~j=1, \ldots k$, $\mu_{0j} = \mathbf{x}^{T}_{j} \mathbf{\beta}$, $x_{j}$ is $j$-th group's covariate vector ($m\times 1$), $m$ is the number of regression coefficients and $\mathbf{\beta}$ and $A$ are unknown. Note that if there is no covariate then $x_{j}=1$ for an intercept term ($m=1$) and so $\mu_{0j}=\mu_{0}=\beta_{0}$ for all $j$, resulting in an exchangeable prior distribution. For reference, a parameter with a zero subscript, such as $\mu_{0j}$, represents a mean parameter of the prior (second-level) distribution, \emph{i.e.}, a prior mean. Based on this conjugate prior distribution it is easy to derive the corresponding posterior distribution
\begin{equation} \label{normalpost}
\mu_{j}\vert \textbf{y}, \beta, A \stackrel{ind}{\sim}\textrm{Normal}(~(1-B_{j})y_{j} + B_{j}\mu_{0j},~(1-B_{j})V_{j}),
\end{equation}
where $B_{j}\equiv V_{j}/(V_{j} + A),~j=1, \ldots, k$, are called shrinkages.

@@

To continue with a maximization based estimation procedure but to steer clear of aforementioned issues we make us of adjustment for density maximization (ADM) \cite{2011}. In general ADM is a procedure to obtain estimates of posterior moments via maximization if the underlying posterior can be approximated by a Pearson family. For our purposes we can approximate the posterior distribution of a shrinkage factor with the Beta distribution allowing us to finally obtain estimates the posterior moments, \emph{i.e.}, $E(B_{j}\vert\textrm{data})$ and $Var(B_{j}\vert\textrm{data})$, without any trouble that MLE can cause. Please refer to \cite{2011} for additional advantages of ADM.

@@

\subsection[Approximation to Posterior Distribution by Moment Matching]{Approximation to Posterior Distribution via Matching Moments}
After estimating the two posterior moments, $E(p_{j}\vert \textbf{z})$ and $Var(p_{j}\vert \textbf{z})$, \code{gbp} reasonably approximates a posterior distribution of the mean effects given the data by assuming a reasonable distribution and matching moments. The Binomial-Beta model we approximate $p_{j}\vert \textbf{z}$ with a Beta distribution and for the Poisson-Gamma model we approximate $\lambda_{j}\vert \textbf{z}$ with a Gamma distribution. Finally for the Normal-Normal model we actually estimate the first three moments and approximate $\mu_{j}\vert \textbf{y}$ with a skewed-t distribution. %TODO: should I put down the form of the skewed-t or just put a reference to it?

@@

There are two kinds of model justification processes; one is a model check and the other is a method check. The model check is for the justification of the model \ref{multilevel} on a specific data set. One possible question is, ``Can this data set benefit from such a multilevel modeling?'' \cite{1997} answered this question by using a mixture model, $z_{j}\vert \beta, r\sim\textrm{Negative-Binomial}$, on Poisson data to justify the second-level hierarchy. They found that their data had more variation than expected of the first-level Poisson distribution and Poisson hierarchical model could successfully account for such additional variation.

Once we are sure that the hierarchical modeling can be appropriate for our data, the following question will be about the validity of interval estimates, the final product of this multilevel modeling. ``Does the 95\% (can be specified differently) confidence interval obtained via this Bayesian model-fitting process achieve 95\% confidence level  for any true parameter values?'' Our answer is ``yes'' and \pkg{Rgbp} has a function to check this point. From now on, all the explanations will be based on the Binomial model because the idea is the same and can be easily applied to the other two models.

\subsection{Pseudo-data Generation Process}
Figure 1 will be helpful to understand this process. As we can see in (8), the distribution of each true batting average ($p_{j},~j=1,\ldots, 18$) depends on two hyper-parameters, $r$ and $\beta_{(m\times1)}$, where $m$ is the number of regression coefficients. So, once we fix these hyper-parameters at specific values, we can generate true batting averages. Suppose we sampled 500 $\mathbf{p}_{(18\times1)}$'s, \emph{i.e.}, $\{\mathbf{p}^{(i)}_{(18\times1)},~i=1, \ldots, 500\}$ from the prior distribution in (8), where $r$ and $\beta$ are given. Then, we can also generate $\{\mathbf{z}^{(i)}_{(18\times1)},~i=1, \ldots, 500\}$ given each $p_{ij},$ where $i$ indicates $i$-th pseudo-data set and $j$ does $j$-th player. Next, \code{coverage} fits the Binomial hierarchical model 500 times on $\{(\mathbf{z}^{(i)}_{(18\times1)}, \mathbf{n}_{(18 \times 1)}),~i=1, \ldots, 500\}$ to obtain 500$\times$18 interval estimates.  After this generating process, as for the first player, we will have 500 pairs of ($z_{i1}$, $p_{i1}$) and 500 interval estimates, $\{(\hat{p}_{i1, low}, ~\hat{p}_{i1, upp}$), $i=1, \ldots, 500\}$.
\begin{figure}[h]
\begin{center}
\includegraphics[width=5cm]{process.png}
\caption{Pseudo-data generating process}
\end{center}
\end{figure}

\subsection{Simple Unbiased Estimates of Coverage Probabilities}
Based on the generated pseudo-data sets, let's define an indicator variable, $I_{ij}$, which is 1 if $j$-th player's interval estimate from $i$-th pseudo-data set includes $p_{ij}$ and 0 otherwise. One way to estimate the coverage probability is taking average over these indicator variables. We call it a simple unbiased coverage probability estimate. For example, $\bar{I}_{1}=\sum_{i=1}^{500}I_{i1}/500$ is the estimated coverage probability for the first player.

\subsection{Rao-Blackwellized Unbiased estimates of Coverage Probabilities}
Rao-Blackwellization improves accuracy of an unbiased estimator by taking conditional expectation given a sufficient statistic. Based on this idea, we define $E(I_{ij}\vert z_{ij}, r, \beta$), where $r$ and $\beta$ were given at first (see 5.1) and $z_{ij}$ is a sufficient statistic. This expectation is the same as Pr$(\hat{p}_{ij, low}\le p_{ij} \le\hat{p}_{ij, upp}\vert z_{ij}, r, \beta)$, where ($\hat{p}_{ij, low}$, $\hat{p}_{ij, upp}$) is $j$-Ch player's interval estimates on the $i$-th data set. We can calculate this probability  because we know the distribution of $p_{ij} \vert z_{ij}, r, \beta$ in (9). Note that conditioning on $z_{ij}$ is equivalent to conditioning on $\mathbf{z}$ as long as $r$ and $\beta$ are known. Then, we can estimate the first player's coverage probability by $\sum_{i=1}^{500}E(I_{i1}\vert z_{i1}, r, \beta)/500$, which is also unbiased but more accurate than the previous simple estimator.

%TODO: Proof read Hospital and Baseball examples
\section[Examples]{Examples}
\subsection[Known Second-level Mean]{31 Hospitals: Known Second-level Mean}
Suppose you live in the New York state (NY) and have been suffering from severe coronary heart disease (hopefully not). If you are supposed to receive the coronary artery bypass graft (CABG) surgery soon, you might want to find the most famous hospital for dealing with such a surgery. On top of that, if you can figure out each hospital's ability to handle this surgery, it will be useful for your decision in choosing a hospital.

@@

As we can see in (6), the posterior mean, $(1-B_{j})y_{j} + B_{j}\lambda_{0}$, is a linear function of shrinkage, $B_{j}\equiv r / (r + n_{j})$, locating between the sample mean and prior mean ($\lambda_{0}=0.02$). It makes sense because $r$ can be interpreted as the amount of prior information and $n_{j}$ as the amount of observed information. If the second level has more information than the first level, then the sample mean shrinks towards the prior mean more than 50\%. This point is clear in the above output because, as caseload increases, shrinkage decreases, depending less on the prior distribution.

@@

  post.mode.alpha post.sd.alpha
1          -5.818         0.411
\end{CodeOutput}
\end{CodeChunk}
Since estimated $\alpha$ is -5.818, we can easily calculate $\hat{r}=\textrm{exp}(5.818)=336$, which is a good indicator of how valuable and informative the hypothetical second-level hierarchy is. As mentioned before, $r$ can be interpreted as the amount of prior information, although we do not know its true value. So, \code{gbp} asked data how much credit they wanted to give to this hypothetical second-level hierarchy. $\hat{r}$ = 336 is the answer from the data. It means that observed sample means of hospitals whose caseloads are less than 336 will shrink toward the prior mean (0.020) more than 50\%.

@@

         obs.mean   se prior.mean shrinkage low.intv post.mean upp.intv post.sd
5           -1.00  9.0      8.168     0.408  -13.297     2.737   16.692   7.634
2            8.00 10.0      8.168     0.459   -7.255     8.077   23.361   7.810
7           18.00 10.0      8.168     0.459   -1.289    13.484   30.821   8.176
4            7.00 11.0      8.168     0.507   -8.780     7.592   23.602   8.257
6            1.00 11.0      8.168     0.507  -13.027     4.633   20.131   8.441
1           28.00 15.0      8.168     0.657   -2.315    14.979   38.763  10.560
3           -3.00 16.0      8.168     0.685  -17.130     4.650   22.477  10.096
8           12.00 18.0      8.168     0.734  -10.208     9.189   29.939  10.227
colMeans     8.75 12.5      8.168     0.552   -9.163     8.168   25.723   8.900
\end{CodeOutput}
\end{CodeChunk}
The output from \code{gbp} provides an easy to read summary of the results of the estimation. From this there seems to be little evidence that the training provided much of an added benefit with every school's $95\%$ posterior interval containing 0. \pkg{Rgbp} also provides functionality to plot the results of the analysis as seen in \ref{8schoolsplot}.

>>>>

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Joseph Kelly, Carl Morris, Hyungsuk Tak} %% comma-separated
\Plaintitle{Rgbp: An R Package for Hierarchical modeling and Method Checking} %% without formatting

@@

where $~j=1, \ldots k$, $\mu_{0j} = x^{T}_{j} \mathbf{\beta}$, $x_{j}$ is the $j$-th group's covariate vector ($m\times 1$), $m$ is the number of regression coefficients and $\mathbf{\beta}$ and $A$ are unknown. Note that if there is no covariate then $x_{j}=1$ for an intercept term ($m=1$) and so $\mu_{0j}=\mu_{0}=\beta_{0}$ for all $j$, resulting in an exchangeable prior distribution. For reference, a parameter with a zero subscript, such as $\mu_{0j}$, represents a mean parameter of the prior (second-level) distribution, \emph{i.e.}, a prior mean. Based on this conjugate prior distribution it is easy to derive the corresponding posterior distribution
\begin{equation} \label{normalpost}
\mu_{j}\vert \textbf{y}, \beta, A \stackrel{ind}{\sim}\textrm{Normal}(~(1-B_{j})y_{j} + B_{j}\mu_{0j},~(1-B_{j})V_{j}),
\end{equation}
where $B_{j}\equiv V_{j}/(V_{j} + A),~j=1, \ldots, k$, are called shrinkages.

@@

To continue with a maximization-based estimation procedure but to steer clear of aforementioned issues we make use of adjustment for density maximization (ADM) \cite{2011}. In general ADM is a procedure to obtain estimates of posterior moments via maximization if the underlying posterior can be approximated by a Pearson family. For our purposes we can approximate the posterior distribution of a shrinkage factor with the Beta distribution allowing us to finally obtain estimates of the posterior moments, \emph{i.e.}, $E(B_{j}\vert\textrm{data})$ and $Var(B_{j}\vert\textrm{data})$, without any trouble that MLE can cause. Please refer to \cite{2011} for additional advantages of ADM.

@@

\subsection[Approximation to Posterior Distribution by Moment Matching]{Approximation to Posterior Distribution via Matching Moments}
After estimating the two posterior moments, $E(p_{j}\vert \textbf{z})$ and $Var(p_{j}\vert \textbf{z})$, \code{gbp} reasonably approximates a posterior distribution of the mean effects given the data by assuming a reasonable distribution and matching moments. For the Binomial-Beta model we approximate $p_{j}\vert \textbf{z}$ with a Beta distribution and for the Poisson-Gamma model we approximate $\lambda_{j}\vert \textbf{z}$ with a Gamma distribution. Finally for the Normal-Normal model we actually estimate the first three moments and approximate $\mu_{j}\vert \textbf{y}$ with a skewed-t distribution. %TODO: should I put down the form of the skewed-t or just put a reference to it?

@@

There are two kinds of model justification processes; one is a model checking and the other is a method checking. The model checking is for the justification of the model \ref{multilevel} on a specific data set. One possible question is, ``Can this data set benefit from such a multilevel modeling?'' \cite{1997} answered this question by using the Negative-Binomial mixture model on Poisson data to justify the second-level hierarchy. They found that their data had more variation than expected of the first-level Poisson distribution and Poisson hierarchical model could successfully account for such additional variation.

Once we are sure that the hierarchical modeling can be appropriate for our data, the following question will be about the validity of interval estimates, the final product of this multilevel modeling. ``Does the 95\% (can be specified differently) confidence interval obtained via this Bayesian model-fitting process achieve 95\% confidence level  for any true parameter values?'' Our answer is ``yes'' and \pkg{Rgbp} has a function to check this point. From now on, all the explanations will be based on the Binomial model using the baseball data because the idea is the same and can be easily applied to the other two models.

\subsection{Pseudo-data Generation Process}
Figure 1 will be helpful to understand this process. As we can see in (8), the distribution of each true batting average ($p_{j},~j=1,\ldots, 18$) depends on two hyper-parameters, $r$ and $\beta_{(2\times1)}$. So, once we fix these hyper-parameters at specific values, we can generate true batting averages. Suppose we sampled 500 $\mathbf{p}_{(18\times1)}$'s, \emph{i.e.}, $\{\mathbf{p}^{(i)}_{(18\times1)},~i=1, \ldots, 500\}$ from the prior distribution in (8), where $r$ and $\beta$ are given. Then, we can also generate $\{\mathbf{z}^{(i)}_{(18\times1)},~i=1, \ldots, 500\}$ given each $p_{ij},$ where $i$ indicates the $i$-th pseudo-data set and $j$ does the $j$-th player. Next, \code{coverage} fits the Binomial hierarchical model 500 times on $\{(\mathbf{z}^{(i)}_{(18\times1)}, \mathbf{n}_{(18 \times 1)}),~i=1, \ldots, 500\}$ to obtain 500$\times$18 interval estimates.  After this generating process, as for the first player, we will have 500 pairs of ($z_{i1}$, $p_{i1}$) and 500 interval estimates, $\{(\hat{p}_{i1, low}, ~\hat{p}_{i1, upp}$), $i=1, \ldots, 500\}$.
\begin{figure}[h]
\begin{center}
\includegraphics[width=5cm]{process.png}
\caption{Pseudo-data generating process}
\end{center}
\end{figure}

\subsection{Simple Unbiased Estimates of Coverage Probabilities}
Based on the generated pseudo-data sets, let's define an indicator variable, $I_{ij}$, which is 1 if the $j$-th player's interval estimate from the $i$-th pseudo-data set includes $p_{ij}$ and 0 otherwise. One way to estimate the coverage probability is taking average over these indicator variables. We call it a simple unbiased coverage probability estimate. For example, $\bar{I}_{1}=\sum_{i=1}^{500}I_{i1}/500$ is the estimated coverage probability for the first player.

\subsection{Rao-Blackwellized Unbiased estimates of Coverage Probabilities}
Rao-Blackwellization improves accuracy of an unbiased estimator by taking conditional expectation given a sufficient statistic. Based on this idea, we define $E(I_{ij}\vert z_{ij}, r, \beta$), where $r$ and $\beta$ were given at first (see 5.1) and $z_{ij}$ is a sufficient statistic. This expectation is the same as Pr$(\hat{p}_{ij, low}\le p_{ij} \le\hat{p}_{ij, upp}\vert z_{ij}, r, \beta)$, where ($\hat{p}_{ij, low}$, $\hat{p}_{ij, upp}$) is the $j$-th player's interval estimates on the $i$-th data set. We can calculate this probability  because we know the distribution of $p_{ij} \vert z_{ij}, r, \beta$ in (9). Note that conditioning on $z_{ij}$ is equivalent to conditioning on $\mathbf{z}$ as long as $r$ and $\beta$ are known. Then, we can estimate the first player's coverage probability by $\sum_{i=1}^{500}E(I_{i1}\vert z_{i1}, r, \beta)/500$, which is also unbiased but more accurate than the previous simple estimator.

%TODO: Proof read Hospital and Baseball examples
\section[Examples]{Examples}
\subsection[Known Second-level Mean]{31 Hospitals: Known Second-level Mean}
Suppose you live in the New York state (NY) and have been suffering from severe coronary heart disease. If you are supposed to receive the coronary artery bypass graft (CABG) surgery soon, you might want to find the most famous hospital for dealing with such a surgery. On top of that, if you can figure out each hospital's ability to handle this surgery, it will be useful for your decision in choosing a hospital.

@@

As we can see in (6), the posterior mean, $(1-B_{j})y_{j} + B_{j}\lambda_{0}$, is a linear function of shrinkage, $B_{j}\equiv r / (r + n_{j})$, locating between the sample mean and prior mean ($\lambda_{0}=0.02$). It makes sense because $r$ can be interpreted as the amount of prior information and $n_{j}$ as the amount of observed information. If the second level has more information than the first level, then the sample mean shrinks towards the prior mean more than 50\%. This point is clear in the above output because, as caseload increases, shrinkage decreases, depending less on the second level information.

@@

  post.mode.alpha post.sd.alpha
1          -5.818         0.411
\end{CodeOutput}
\end{CodeChunk}
Since estimated $\alpha$ is -5.818, we can easily calculate $\hat{r}=\textrm{exp}(5.818)=336$, which is a good indicator of how valuable and informative the hypothetical second-level hierarchy is. As mentioned before, $r$ can be interpreted as the amount of prior information, although we do not know its true value. So, \code{gbp} asked the data how much credit they wanted to give to this hypothetical second-level hierarchy and $\hat{r}$ = 336 was their answer. It means that observed sample means of hospitals whose caseloads are less than 336 will shrink toward the prior mean (0.020) more than 50\%.

@@

         obs.mean   se prior.mean shrinkage low.intv post.mean upp.intv post.sd
5           -1.00  9.0      8.168     0.408  -13.297     2.737   16.692   7.634
2            8.00 10.0      8.168     0.459   -7.255     8.077   23.361   7.810
7           18.00 10.0      8.168     0.459   -1.289    13.484   30.821   8.176
4            7.00 11.0      8.168     0.507   -8.780     7.592   23.602   8.257
6            1.00 11.0      8.168     0.507  -13.027     4.633   20.131   8.441
1           28.00 15.0      8.168     0.657   -2.315    14.979   38.763  10.560
3           -3.00 16.0      8.168     0.685  -17.130     4.650   22.477  10.096
8           12.00 18.0      8.168     0.734  -10.208     9.189   29.939  10.227
colMeans     8.75 12.5      8.168     0.552   -9.163     8.168   25.723   8.900
\end{CodeOutput}
\end{CodeChunk}
The output from \code{gbp} provides an easy way to read summary of the results of the estimation. From this there seems to be little evidence that the training provided much of an added benefit with every school's $95\%$ posterior interval containing 0. \pkg{Rgbp} also provides functionality to plot the results of the analysis as seen in \ref{8schoolsplot}.
